---
title: "Using `dada2` to process AGP 16S rRNA data"
author: "Quang Nguyen"
date: "2022-03-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", autodep = TRUE)
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::knit_engines$set(targets = targets::tar_engine_knitr)
}
```

Here, we use `dada2` to pre-process all data from AGP and Gevers et al.

```{r basic_load}
library(dada2)
library(here)
library(targets)
library(tidyverse)
library(tarchetypes)

Sys.setenv(TAR_PROJECT = "dada2_agp")
tar_unscript()
here::i_am("analysis/agp_dada2.Rmd")
base_path <- file.path("rc", "lab", "H", 
                          "HoenA", "Lab", "QNguyen", "ResultsFiles", "data", 
                          "agp")
if (Sys.info()['sysname'] == "Darwin"){
    filepath <- file.path("/Volumes", base_path)
    if (!file.exists(file.path(filepath, ".placeholder"))){
        stop("This folder does not exist, please mount folder")
    } 
} else if (Sys.info()['sysname'] == "Linux"){
    filepath <- file.path("/dartfs-hpc", base_path)
} else {
    stop("OS not supported")
}
```

First, let's visualize the quality profiles of some sample reads

```{r}
reads <- sort(list.files(file.path(filepath, "raw"), pattern = "_001.fastq", full.names = TRUE))[1:10]
sample_names <- sapply(str_split(basename(reads), "_001.fastq"), `[`, 1)
suppressWarnings(plotQualityProfile(reads))
```

We can see that the sequencing runs are actually pretty good, so we're going to trim off the last 10 nucleotides as suggested in the [`dada2` tutorial](https://benjjneb.github.io/dada2/tutorial.html). Since AGP data is not paired end data, we skip the merging step and having to pre-process each forward and reverse read separately.

```{r}
filt <- file.path(filepath, "filtered", paste0(sample_names, "_filt.fastq.gz"))
trim <- filterAndTrim(reads, filt, truncLen = 145, maxEE=1, truncQ=11, rm.phix = TRUE, compress=TRUE, verbose=TRUE, multithread = TRUE)
err <- learnErrors(filt, multithread = TRUE)
plotErrors(err, nominalQ = TRUE)
```

We iterate through a basic set of samples and have the pipeline run through the entire 18,000 samples on the cluster. This is done through the usage of the `targets` package. The pipeline definition script can be found at `script_dada2_agp.R`  

```{r}
library(targets)
Sys.setenv(TAR_PROJECT = "dada2_agp")
targets::tar_visnetwork(targets_only = TRUE)
```

